Aspects of Dealing with Imperfect Data in Temporal Databases

### 2012-03-16 trunk/book/sources/timeDomain.tex
And (hopefully) the last commit. We're as good as done.

Data models used by information systems (and in specific, temporal database systems) may represent these underlying time domains using \emph{chronons}~\cite{Dyreson1994}. However, to explain what chronons are, an explanation of some other temporal concepts is necessary.

@@

Orthogonal to the classification of underlying time models as linear or cyclic, they can be classified as \emph{discrete}, \emph{dense} or \emph{continuous} models\cite{Dyreson1994}. In a discrete model~\cite{Clifford:1985:AHR:971699.318922}, the notion exists that every instant has a unique successor and the set of instants is seen as a discrete one. Here, intuitively, the set of instants can be seen as isomorphic to the set of natural numbers $\Nat$. In a dense model, the notion exists that between any two instants always lies another. Here, intuitively, the set of instants can be seen as isomporphic to the set of rational numbers $\Q$ (when the set of instants is a discrete one) or the set of real numbers $\R$ (when the set of instants is a continuous one). In a continuous model, the notion also exists that between any two instants always lies another one, but the set of instants is always seen as continuous and there are no ``gaps'' between successive instants.

Some other necessary concepts are ~\cite{Dyreson1994}:

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Time Interval}\\
A \emph{\textbf{time interval}} is the time between two instants.
\end{definition}

\begin{definition}\textbf{Temporal Element}\\
A \emph{\textbf{temporal element}} is a finite union of time intervals.
\end{definition}

\begin{definition}\textbf{Event}\\
An \emph{\textbf{event}} is an instantaneous fact, i.e. something occurring at an instant.
\end{definition}

\begin{definition}\textbf{Duration}\\
A \emph{\textbf{duration}} is an amount of time with known length, but no specific starting or ending instants.
\end{definition}

\begin{definition}\textbf{`Temporal' as Modifier}\\
The modifier \emph{\textbf{`temporal'}} is used to indicate that the modified concept concerns some aspect of time.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

Data models might now represent an underlying time domain using chronons:

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Chronon}\\
In a data model, a \emph{\textbf{chronon}} is a non-decomposable time interval of some fixed, minimal duration.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

Another classification of time models concerns the use of points or intervals to model time. The equivalence between interval-based and point-based time models is demonstrated in \cite{Böhlen_point-versus}.

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Finner Than}\\
Consider a mapping function $f$ and let $i_G$ and $j_H$ be elements of granularities $G$ and $H$ respectively. Granularity $G$ is now said to be \emph{\textbf{finner than}} granularity $H$ if:
\begin{equation}
\label{eq:finner-than}
| f \left( i_G \right) |\ <\ |  f \left( j_H \right)| \nonumber
\end{equation} 
\end{definition}

\begin{definition}
\label{def:coarser-than}\textbf{Coarser Than}
Consider a mapping function $f$ and let $i_G$ and $j_H$ be elements of granularities $G$ and $H$ respectively. Granularity $G$ is now said to be \emph{\textbf{coarser than}} granularity $H$ if:
\begin{equation}
\label{eq:coarser-than}
| f \left( i_G \right)|\ >\ |  f \left( j_H \right)| \nonumber
\end{equation} 
\end{definition}
\vspace{-10pt}
\end{svgraybox}

It is also possible to describe the relation between different granularities. This is called a casting function:
\begin{svgraybox}
\begin{definition}
\textbf{Casting function}~\cite{Lin97}\\
\label{def:casting-fucntion}
Consider two different granularities $G$ and $H$. A granularity-to-granularity \emph{\textbf{casting function}} $cast$ is then a function mapping granules from $G$ to granules from $H$: 
\begin{align}
\label{eq:cast-function}
%\mbox{cast} \left( i_G,G,H \right) = j_H \nonumber
cast & : G \times \mathbb{G} \times \mathbb{G} \rightarrow H \nonumber\\
		 & : (i_{G}, G, H) \rightarrow j_{H} \nonumber
\end{align}
where $i_{G} \in G$ and $j_{H} \in H$ and where $\mathbb{G}$ denotes the set of all granularities.
\end{definition}
\end{svgraybox}

Thus, the function $cast$ associates a granule $i_G$ in $G$ to a corresponding granule $j_H$ in $H$. Two kinds of granularity-to-granularity mappings can now be discerned: an \emph{upwards mapping} is a mapping from a granularity $G$ to a coarser granularity $H$, whereas a downwards mapping is a mapping from a granularity $K$ to a finner granularity $L$. Different granularity-to-granularity mappings between several granularities can be represented using a granularity graph, which is a directed graph indicating the mapping conversions.

Mappings between two granularities may be classified as \emph{regular mappings}, \emph{irregular mappings} or \emph{congruent mappings}~\cite{Lin97},~\cite{DyresonSnodgrass1994}.
\begin{itemize}
\item
\emph{Regular mapping}: A regular mapping is a granularity-to-granularity mapping where the mapping function value is calculated by means of multiplications and/or divisions and (maybe) an anchor adjustment. E.g. the mapping value of the mapping between hours and minutes is calculated using a multiplication by $60$. 
\item
\emph{Irregular mapping}: An irregular mapping is a granularity-to-granularity mapping where the mapping function value can not be calculated by means of multiplications and/or divisions. E.g. the mapping value of the mapping between months and days is dependent on the exact month or day. 
\item
\emph{Congruent mapping}: A congruent mapping is a granularity-to-granularity mapping where the two granularities involved in the mapping have the same granules but a different anchor. E.g. the mapping between the days (Gregorian calendar days) and the academic days is a congruent mapping.
\end{itemize}

The above is illustrated in the following example.

\begin{example}
Consider a system that models both Gregorian Calendar dates as well as Academic dates. The chronons granularity in this system contains milliseconds. Figure \ref{fig:granularity-graph-example} shows the complete granularity graph corresponding with this example. The transitions between Chronons and Seconds and between Seconds and Minutes and so on is called a \emph{regular mapping}. The transition between Days and Months is not defined by a multiplication or division, this is called a \emph{irregular mapping}. In the graph, these \emph{irregular mappings} are noted with a bold arrow. Finally, the transition between a Gregorian Calendar Day and an Academic Day is called a \emph{Congruent mapping}: two granularities with the same granules but with different anchors. Both concern the same days, but the academic year starts on October 1st, whereas the Gregorian calendar year starts on January 1st. Congruent mappings are visualised as straight lines without arrow heads.

@@

\begin{figure}
\centering
\includegraphics[scale=0.5]{graphs/granularityGraph.eps}
\caption{A granularity graph example}
\label{fig:granularity-graph-example}
\end{figure}

Next, in subsection \ref{subsec:timedomain-calendar}, some basic concepts and issues concerning calendars are explained.

\subsection{Calendars}
\label{subsec:timedomain-calendar}
%maybe introduce previously the concept of time units, linear hierarchy and time point.

This subsection will concern \emph{calendars}. A calendar provides some sort of human interpretation of time. As such, calendars provide meaning or interpretation to temporal values where this meaning or interpretation is relevant or useful to the user. More precisely, calendars determine the mapping between human-meaningful time values and an underlying time model~\cite{Dyreson1994}. More formally, a calendar is an organization of different granularities (e.g. granularities day, month and year) in a hierarchy. In order to formally define what a calendar is, some other concepts should first be introduced~\cite{Kraus1997}:

@@

The next section briefly introduces temporal relationships.

\subsection{Temporal Relationships}
%an explanation on allen's temporal relations.
%definition of fuzzy temporal interval
In this section, a brief introduction can be found, concerning \emph{temporal relationships}, sometimes also called `temporal relations'\cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy}. Temporal relationships can be seen as relationships between temporal elements or instants.

\def\JPicScale{0.5}
\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{graphs/allen.eps}
\caption{Allen relations between two time intervals.}
\label{fig:allen}
\end{figure}

Several operators are defined in order to compare temporal elements or time points. Allen~\cite{Allen83} first described the relationships between time intervals and as a special case, between time points. Figure \ref{fig:allen} shows the temporal relationships Allen discerned. Some proposals can be applied to both crisp and fuzzy temporal intervals~\cite{ohlbach2004},~\cite{nagypal2003},~\cite{schockaert08},~\cite{garrido2009}.

In the next section, some basic concepts and terminology about temporal databases are presented, followed by an overview of some interesting issues concerning temporal databases and a survey of some commercial temporal database systems.

>>>>

Data models used by information systems (and in specific, temporal database systems) may represent these underlying time domains using \emph{chronons}~\cite{Dyreson1994}, which can informally be described as the smallest distinguishable time units that can be used in the system. However, to explain what chronons are, an explanation of some other temporal concepts is necessary.

@@

Orthogonal to the classification of underlying time models as linear or cyclic, they can be classified as \emph{discrete}, \emph{dense} or \emph{continuous} models\cite{Dyreson1994}. In a discrete model~\cite{Clifford:1985:AHR:971699.318922}, the notion exists that every instant has a unique successor and the set of instants is seen as a discrete one. Here, intuitively, the set of instants can be seen as isomorphic to the set of natural numbers $\Nat$. In a dense model, the notion exists that between any two instants always lies another. Here, intuitively, the set of instants can be seen as isomorphic to the set of rational numbers $\Q$ (when the set of instants is a discrete one) or the set of real numbers $\R$ (when the set of instants is a continuous one). In a continuous model, the notion also exists that between any two instants always lies another one, but the set of instants is always seen as continuous and there are no ``gaps'' between successive instants.

Some other necessary concepts are:

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Time Interval}~\cite{Dyreson1994}\\
A \emph{\textbf{time interval}} is the time between two instants.
\end{definition}

\begin{definition}\textbf{Duration}~\cite{Dyreson1994}\\
A \emph{\textbf{duration}} is an amount of time with known length, but no specific starting or ending instants.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

A time interval as such is bounded by two instants, whereas a duration is not. Also, it should be noted that an instant is in fact a singular case of a time interval.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Temporal Element}~\cite{Dyreson1994}\\
A \emph{\textbf{temporal element}} is a finite union of time intervals.
\end{definition}

\begin{definition}\textbf{Event}~\cite{Dyreson1994}\\
An \emph{\textbf{event}} is an instantaneous fact, i.e. something occurring at an instant.
\end{definition}

\begin{definition}\textbf{`Temporal' as Modifier}~\cite{Dyreson1994}\\
The modifier \emph{\textbf{`temporal'}} is used to indicate that the modified concept concerns some aspect of time.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

Data models for time modelling might now represent an underlying time domain using chronons:

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Chronon}~\cite{Dyreson1994}\\
In a data model, a \emph{\textbf{chronon}} is a non-decomposable time interval of some fixed, minimal duration.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

@@

Another classification of time models concerns the use of points or intervals to model time. The equivalence between interval-based and point-based time models is demonstrated in~\cite{Bohlen1998}.

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Finner Than}~\cite{Lin97}\\
Consider a mapping function $f$ and let $i_G$ and $j_H$ be elements of granularities $G$ and $H$ respectively. Granularity $G$ is now said to be \emph{\textbf{finner than}} granularity $H$ if:
\begin{equation}
\label{eq:finner-than}
| f \left( i_G \right) |\ <\ |  f \left( j_H \right)| \nonumber
\end{equation} 
\end{definition}

\begin{definition}
\label{def:coarser-than}\textbf{Coarser Than}~\cite{Lin97}\\
Consider a mapping function $f$ and let $i_G$ and $j_H$ be elements of granularities $G$ and $H$ respectively. Granularity $G$ is now said to be \emph{\textbf{coarser than}} granularity $H$ if:
\begin{equation}
\label{eq:coarser-than}
| f \left( i_G \right)|\ >\ |  f \left( j_H \right)| \nonumber
\end{equation} 
\end{definition}
\vspace{-10pt}
\end{svgraybox}

It is also possible to describe the relation between different granularities. This is called a casting function:
\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
\textbf{Casting function}~\cite{Lin97}\\
\label{def:casting-fucntion}
Consider two different granularities $G$ and $H$. A granularity-to-granularity \emph{\textbf{casting function}} $cast$ is then a function mapping granules from $G$ to granules from $H$: 
\begin{align}
\label{eq:cast-function}
%\mbox{cast} \left( i_G,G,H \right) = j_H \nonumber
cast & : G \times \mathbb{G} \times \mathbb{G} \rightarrow H \nonumber\\
		 & : (i_{G}, G, H) \rightarrow j_{H} \nonumber
\end{align}
where $i_{G} \in G$ and $j_{H} \in H$ and where $\mathbb{G}$ denotes the set of all granularities.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

Thus, the function $cast$ associates a granule $i_G$ in $G$ to a corresponding granule $j_H$ in $H$. Two kinds of granularity-to-granularity mappings can now be discerned: an \emph{upwards mapping} is a mapping from a granularity $G$ to a coarser granularity $H$, whereas a downwards mapping is a mapping from a granularity $K$ to a finner granularity $L$. Orthogonal to this classification, mappings between two granularities may be classified as \emph{regular mappings}, \emph{irregular mappings} or \emph{congruent mappings}~\cite{Lin97},~\cite{DyresonSnodgrass1994}.
\begin{itemize}
\item
\emph{Regular mapping}: A regular mapping is a granularity-to-granularity mapping where the mapping function value is calculated by means of multiplications and/or divisions and (maybe) an anchor adjustment. e.g., the mapping value of the mapping between hours and minutes is calculated using a multiplication by $60$. 
\item
\emph{Irregular mapping}: An irregular mapping is a granularity-to-granularity mapping where the mapping function value can not be calculated by means of multiplications and/or divisions. e.g., the mapping value of the mapping between months and days is dependent on the exact month or day. 
\item
\emph{Congruent mapping}: A congruent mapping is a granularity-to-granularity mapping where the two granularities involved in the mapping have the same granules but a different anchor. e.g., the mapping between the days (Gregorian calendar days) and the academic days is a congruent mapping.
\end{itemize}

Different granularity-to-granularity mappings between several granularities can be represented using a granularity graph, which is a directed graph indicating the mapping conversions. The above is illustrated in the following example.

\begin{example}
\label{examplegrangraph}
Consider a system that models both Gregorian calendar dates as well as academic calendar dates. In this system, the chronons granularity is a set of milliseconds. Figure~\ref{fig:granularity-graph-example} shows the complete granularity graph corresponding to this example. The transition between the chronons granularity and the seconds granularity is an example of a regular mapping. Regular mappings are represented by thin arrows in the visualisation of the graph. The transition between the days granularity and the months granularity is an example of an irregular mapping. In the graph visualisation, irregular mappings are represented by a bold arrow. Finally, the transition between the Gregorian calendar day granularity and the academic day granularity is an example of a congruent mapping. Both concern the same days, but the academic year starts on October 1st, whereas the Gregorian calendar year starts on January 1st. In the graph visualisation, congruent mappings are visualised as straight lines without arrow heads.

@@

\begin{figure}
\centering
\includegraphics[scale=0.5]{graphs/granularityGraph.eps}
\caption{The granularity graph corresponding to example~\ref{examplegrangraph}.}
\label{fig:granularity-graph-example}
\end{figure}

\vspace{-25pt}

%Next, in subsection \ref{subsec:timedomain-calendar}, some basic concepts and issues concerning calendars are explained.

%\subsection{Calendars}
%\label{subsec:timedomain-calendar}
%maybe introduce previously the concept of time units, linear hierarchy and time point.

%This subsection will concern \emph{calendars}. A calendar provides some sort of human interpretation of time. As such, calendars provide meaning or interpretation to temporal values where this meaning or interpretation is relevant or useful to the user. More precisely, calendars determine the mapping between human-meaningful time values and an underlying time model~\cite{Dyreson1994}. More formally, a calendar is an organization of different granularities (e.g., granularities day, month and year) in a hierarchy. In order to formally define what a calendar is, some other concepts should first be introduced~\cite{Kraus1997}:

@@

%The next section briefly introduces temporal relationships.

\subsection{\label{allen-crisp}Temporal Relationships}
%an explanation on allen's temporal relations.
%definition of fuzzy temporal interval
In this section, a brief introduction can be found, concerning \emph{temporal relationships}, sometimes also called `temporal relations'\cite{Billiet:Pons:Matthe:DeTre:Pons:2011:BipolarFuzzy}. Temporal relationships can be seen as relationships between temporal elements belonging to the same time domain. These relationships express how the temporal elements are related to one another, with respect to temporal precedence and overlap.

\def\JPicScale{0.5}
\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{graphs/allen.eps}
\caption{Allen relations between two time intervals I and J.}
\label{fig:allen}
\end{figure}

Several (collections of) operators have been proposed in order to compare temporal elements and model the temporal relationships between them. Allen~\cite{Allen83} most notably described such relationships between time intervals and as a special case, between instants. Figure~\ref{fig:allen} shows the temporal relationships Allen discerned. Some proposals can be applied to both crisp and other time intervals~\cite{ohlbach2004},~\cite{nagypal2003},~\cite{schockaert08},~\cite{garrido2009}.

%In the next section, some basic concepts and terminology about temporal databases are presented, followed by an overview of some interesting issues concerning temporal databases and a survey of some commercial temporal database systems.
